My title

### 2024-10-31 paper/paper.qmd
citations and justifications in model part done

Polling data for the 2024 U.S. presidential election is provided by FiveThirtyEight [@fivethirtyeight]. This dataset compiles information from numerous polls conducted by various pollsters, detailing public support levels for presidential candidates. This data is periodically updated on FiveThirtyEight's website to reflect the latest polling information for the 2024 election. Each record includes information we need like state in which the poll was conducted (or conducted nationally), polling dates (start and end), the specific pollster, the sample size of each poll, the polling methodology, and support percentages for each candidate. This dataset serves as the foundation for analyzing trends and building predictive models for the winner of the upcoming US presidential election. Table tbl-pollingdata below previews this information for polls conducted in 2023 and 2024.

@@

### Outcome Variables
In this analysis of Donald Trump’s support rate in the 2024 U.S. presidential election, the outcome variable is Trump’s support rate (pct) because it is directly related to the election results. The primary research goal of this analysis is to predict Trump’s chances of winning in the 2024 election, so the support rate is the main metric we aim to model and forecast. The support rate here represents the proportion of respondents in a specific poll who support Trump. We seek to understand how the support rate is influenced by different factors, such as polling methodology, time, or geographical variables. By modeling the fluctuations in Trump’s support rate, we can infer his electoral performance, and assess the impact of various factors on his popularity. Understanding these dynamics provides insights into Trump’s standing relative to his opponents and supports our overall election prediction.

@@

**methodology**: Different survey methods (e.g., telephone interviews, online questionnaires) significantly impact the representativeness of the results, as they directly influence the characteristics of the sampled population. Some methods may introduce biases toward specific demographic groups.

@@

The plot on the left side in @fig-pct1 shows the average support rates for the top and bottom ten pollsters, highlighting each pollster’s most commonly used methodology. We can observe that the support rates between the top 5 and bottom 5 pollsters differ by about 10%, which helps us understand the bias introduced by certain pollsters’ influence on support rates, as some pollsters may tend to show higher or lower support rates. This chart also examines the different methodologies used by these pollsters, as varying survey methods (e.g., phone surveys or online panels) may affect the results by reaching different groups of voters. Focusing on these differences aids in assessing the reliability and bias of the data.

@@

# Model
Our model applies a hierarchical Bayesian approach, based on the framework proposed by Gelman et al. (2020).   This method builds a layered structure that combines polling data collected by different organizations using various methodologies with regional and temporal trends to more accurately predict support rates on Election Day.   Running within a Bayesian framework, our model employs a random walk prior to reflect the dynamically changing support rates following Biden's withdrawal in 2024.   This prior distribution is based on historical election trends and the latest polling data, allowing the model to continuously update as new data becomes available.

The goal of our modelling strategy is to build a time series model to predict the percentage of support for Trump (pct) over time. This model incorporates multiple variables, including the polling methodology (methodology), the number of days since Biden's withdrawal from the race on 21 July 2024 (days_since_Biden_Withdrawal), whether the poll is at the national level (is_national), as well as random effects for each state (state) and each pollster (pollster).  These variables are chosen to capture the effects of time, polling characteristics, and geographic location on support rates, allowing us to obtain dynamic estimates of support for Trump and to predict the winner of 2024 U.S. presidential election.

@@

## Model justification
### Model Prior Selection, Assumptions, and Validation
We based our prior choices on historical polling data and previous research in electoral studies, aiming to provide informative priors without overly constraining the model. For instance, we set a prior on the intercept as $\text{Normal} \sim (45, 5)$, assuming Trump’s support rate centers around 45%, with some uncertainty. Effect size priors (e.g., $\text{Normal} \sim (0, 5)$ for variables such as methodology and days_since_Biden_Withdrawal) reflect our expectation of moderate effects for these variables. This balance between informative priors and adaptability allows the model to accommodate data while providing a reasonable starting framework.

@@

We implemented the model using the rstanarm package, leveraging MCMC sampling for Bayesian inference. To validate the model, we performed out-of-sample testing and calculated the RMSE (Root Mean Square Error), which resulted in 2.31. In the context of complex, socially influenced political surveys, this error level is considered acceptable. Convergence diagnostics show that R-hat values are close to 1, and the Effective Sample Size (ESS) is generally large (often over 400), indicating good independence among samples. Sensitivity analyses on priors indicate that slight adjustments do not significantly alter the predictions, demonstrating model stability.

We also considered a simpler linear regression model (excluding hierarchical structures) and a more complex model (incorporating additional economic data like state GDP and personal income). Through testing, we found that the simpler model could not adequately capture differences across polling agencies and states, while the more complex model added computational cost with minimal gains in accuracy. Consequently, we chose the current approach to balance flexibility, interpretability, and computational efficiency.

@@

For the variable `days_since_Biden_Withdrawal`, we hypothesize that, over time, Trump’s approval ratings may fluctuate due to shifts in the election landscape, policy discussions, and changes in voter sentiment. This hypothesis is based on the dynamic nature of polling data: as elections near their end, poll results typically converge, reflecting more accurate forecasts for election day (see Blumenthal's Polls, Forecasts, and Aggregators). We anticipate that the `days_since_Biden_Withdrawal` variable may exhibit a positive or negative trend, depending on the political climate and voter responses to Trump in particular periods.

Through the `is_national` variable, we distinguish between national and state-level polls, this variable enables the model to capture support rate differences arising from geographic and demographic differences. We expect that national polls, which typically have larger and more representative samples, are better suited to capturing overall trends, while state-level polls may reflect finer, localized biases and specific regional issues because national polls often encompass a broader and more balanced voter base, they provide a more stable forecast of overall voter trends. In contrast, due to geographic and demographic variations, state-level polls often exhibit greater fluctuations in error than national polls. These fluctuations stem from regional issues and diverse demographic characteristics, which may not align fully with national trends (Disentangling Bias and Variance in Election Polls).

>>>>

Polling data for the 2024 U.S. presidential election is provided by @fivethirtyeight. This dataset compiles information from numerous polls conducted by various pollsters, detailing public support levels for presidential candidates. This data is periodically updated on FiveThirtyEight's website to reflect the latest polling information for the 2024 election. Each record includes information we need like state in which the poll was conducted (or conducted nationally), polling dates (start and end), the specific pollster, the sample size of each poll, the polling methodology, and support percentages for each candidate. This dataset serves as the foundation for analyzing trends and building predictive models for the winner of the upcoming US presidential election. Table tbl-pollingdata below previews this information for polls conducted in 2023 and 2024.

@@

### Outcome Variables
In this analysis of Donald Trump’s support rate in the 2024 U.S. presidential election, the outcome variable is Trump’s support rate (i.e., pct) because it is directly related to the election results. The primary research goal of this analysis is to predict Trump’s chances of winning in the 2024 election, so the support rate is the main metric we aim to model and forecast. The support rate here represents the proportion of respondents in a specific poll who support Trump. We seek to understand how the support rate is influenced by different factors, such as polling methodology, time, or geographical variables. By modeling the fluctuations in Trump’s support rate, we can infer his electoral performance, and assess the impact of various factors on his popularity. Understanding these dynamics provides insights into Trump’s standing relative to his opponents and supports our overall election prediction.

@@

**methodology**: We anticipate different survey methods significantly impact the representativeness of the results, as they directly influence the characteristics of the sampled population. Some methods may introduce biases toward specific demographic groups.

@@

The plot on the left side in @fig-pct1 shows the average support rates for the top and bottom ten pollsters, highlighting each pollster’s most commonly used methodology. We can observe that the support rates between the top 5 and bottom 5 pollsters differ by about 10%, which helps us understand the bias introduced by certain pollsters’ influence on support rates, as some pollsters may tend to show higher or lower support rates. This chart also examines the different methodologies used by these pollsters, as varying survey methods may affect the results by reaching different groups of voters. Focusing on these differences aids in assessing the reliability and bias of the data.

@@

# Model
Our model applies a hierarchical Bayesian approach, based on the framework proposed by @gelman2020regression. This method builds a layered structure that combines polling data collected by different organizations using various methodologies with regional and temporal trends to more accurately predict support rates on Election Day. Running within a Bayesian framework, our model employs a random walk prior to reflect the dynamically changing support rates following Biden's withdrawal in 2024. This prior distribution is based on historical election trends and the latest polling data, allowing the model to continuously update as new data becomes available.

The goal of our modelling strategy is to build a time series model to predict the percentage of support for Trump over time. This model incorporates multiple variables, including the polling methodology (methodology), the number of days since Biden's withdrawal from the race on 21 July 2024 (days_since_Biden_Withdrawal), whether the poll is at the national level (is_national), as well as random effects for each state (state) and each pollster (pollster).  These variables are chosen to capture the effects of time, polling characteristics, and geographic location on support rates, allowing us to obtain dynamic estimates of support for Trump and to predict the winner of 2024 U.S. presidential election.

@@

## Model justification
### Model Prior Selection, Assumptions, and Validation
We based our prior choices on historical polling data and previous research in electoral studies, aiming to provide informative priors without overly constraining the model. For instance, we set a prior on the intercept as $\text{Normal} \sim (45, 5)$, assuming Trump’s support rate centers around 45%, with some uncertainty. Effect size priors (e.g., $\text{Normal} \sim (0, 5)$ for variables such as `methodology`, `days_since_Biden_Withdrawal` and `is_national`) reflect our expectation of moderate effects for these variables. This balance between informative priors and adaptability allows the model to accommodate data while providing a reasonable starting framework.

@@

We implemented the model using the rstanarm package, leveraging MCMC sampling for Bayesian inference. To validate the model, we performed out-of-sample testing and calculated the RMSE (Root Mean Square Error), which resulted in 2.31. In the context of complex, socially influenced political surveys, this error level is considered acceptable. Convergence diagnostics show that R-hat values are close to 1, and the Effective Sample Size (ESS) is generally large (often > 400), indicating good independence among samples. Sensitivity analyses on priors indicate that slight adjustments do not significantly alter the predictions, demonstrating model stability.

We also considered a simpler linear regression model which excluding hierarchical structures and a more complex model which incorporating additional economic data like state GDP and personal income. Through testing, we found that the simpler model could not adequately capture differences across polling agencies and states, while the more complex model added computational cost with minimal gains in accuracy. Consequently, we chose the current approach to balance flexibility, interpretability, and computational efficiency.

@@

For the variable `days_since_Biden_Withdrawal`, we hypothesize that, over time, Trump’s approval ratings may fluctuate due to shifts in the election landscape, policy discussions, and changes in voter sentiment. This hypothesis is based on the dynamic nature of polling data: as elections near their end, poll results typically converge, reflecting more accurate forecasts for election day (see @blumenthal2010polls). We anticipate that the `days_since_Biden_Withdrawal` variable may exhibit a positive or negative trend, depending on the political climate and voter responses to Trump in particular periods.

Through the `is_national` variable, we distinguish between national and state-level polls, this variable enables the model to capture support rate differences arising from geographic and demographic differences. We expect that national polls, which typically have larger and more representative samples, are better suited to capturing overall trends, while state-level polls may reflect finer, localized biases and specific regional issues because national polls often encompass a broader and more balanced voter base, they provide a more stable forecast of overall voter trends. In contrast, due to geographic and demographic variations, state-level polls often exhibit greater fluctuations in error than national polls. These fluctuations stem from regional issues and diverse demographic characteristics, which may not align fully with national trends (see @kennedy2018disentangling).
